---
title: |
    | "Project_Pegasus"
    | "An Analysis Differentiating Members from Casual Riders" 
author: "Brian Rice"
date: "2023-08-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library("tidyverse")
library("ggplot2")
library("skimr")
library("dplyr")


bike <- read_csv(
  "C:\\Users\\brice\\Downloads\\202004-divvy-tripdata\\202004-divvy-tripdata.csv")
```

```{r examine, include=TRUE}
skim_without_charts(bike)
```
## Overview
84,776 records
13 fields
dtypes: char, num, POSIXct

Skimming we notice there are 99 end station records missing,
which correspond to 99 end_lat and end_lng records which are also missing.

```{r missing_data, include=FALSE}
bike_missing <- bike[is.na(bike$end_station_name),]

#I do not discern any patterns in the missing data
#since 99 of 84,776 total records is only slightly more 
#than .1% of the total data we'll drop these records
```

Lacking instructions we will drop these records from our analysis.
We will also drop columns which do not bear on our analysis,
like the station_names, rideable type (which has the same value 
for every record).

```{r drop_missing, include=TRUE}

clean_bike <- na.omit(bike)
clean_bike <- clean_bike[, -c(2, 5, 7)]
```

## Adding Columns: 
* ride_duration
* start_time
* start_hour
* start_day
* part_of_day

```{r adding_cols, include=TRUE}
clean_bike$ride_duration <- clean_bike$ended_at - clean_bike$started_at

clean_bike$start_time <- format(clean_bike$started_at, format = '%H:%M:%S')
clean_bike$start_hour <- format(clean_bike$started_at, format = '%H')

clean_bike$start_day <- weekdays(clean_bike$started_at)

clean_bike <- clean_bike %>% 
  mutate(part_of_day = case_when(
    start_hour >= 5 & start_hour < 10 ~ 'morning',
    start_hour >= 10 & start_hour < 16 ~ 'mid',
    start_hour >= 16 & start_hour < 20 ~ 'evening',
    start_hour >= 20 | start_hour < 5 ~ 'late'
  ))
```

## Clean and Drop
* filter out negative ride_durations as obvious errors
* filter out ride_duration = 0 as probably user errors
* dropping columns we no longer need:
  + ride_id
  + started_at
  + ended_at
  + start_time

```{r clean_and_drop, include=TRUE}

clean_bike <- clean_bike %>% 
  filter(ride_duration > 0)

clean_bike <- clean_bike[, -c(1, 2, 3, 12)]
```
## First Plot

```{r plot_first, include=TRUE}
ggplot(data = clean_bike) +
  geom_bar(mapping = aes(x = start_hour)) +
  facet_wrap(~member_casual)
```

## First Impression

We can discard the hours of midnight to 5am as there are virtually no users
between those times.

```{r 5am, include=FALSE}
clean_bike$start_hour <- as.numeric(clean_bike$start_hour)

late_crowd  <- clean_bike %>% 
  filter(start_hour < 5)
ggplot(data = clean_bike) +
  geom_bar(data = subset(clean_bike, start_hour < 5), 
           mapping = aes(x = start_hour))

count(late_crowd)
count(late_crowd) / count(clean_bike) * 100

#the total number of customers taking rides between midnight and 5am are 879.
#This is about 1% of our total so we'll drop these hours to focus more tightly
#on the busier times

clean_bike <- clean_bike %>% 
  filter(start_hour >= 5)
```
## Improved Plot
Having removed the midnight to 5am time range our plot looks much better.

Here we can see the member users significantly outnumber the casual riders,
and while the members see the real start of business around 6am it isn't 
until 11am that the casual riders even begin to match that. 

The casual riders peak at about 2,800 at 2pm, remain steady for several 
more hours, before declining. 
Whereas the members peak at over 6,000 at 5pm and decline rapidly thereafter.


```{r plot_improved, echo=FALSE}
ggplot(data = clean_bike) +
  geom_bar(mapping = aes(x = start_hour, fill='orange')) +
  facet_wrap(~member_casual)
```

## Analysis by Day

These views are fascinating as they reveal that while the casual riders favor
Saturday and Sunday rides (as one might expect), the members also have a
strong usage profile on the weekend.
This is important because it suggests that the member riders are not simply
using the bikes as a means of transportation but might also be using the bikes 
for recreation.
Otherwise the profiles fit the more generalized view of casual riders
favoring later in the day while members tend to be busiest around the end
of the workday.

```{r plot_analysis_by_day, echo=FALSE}

ggplot(data = clean_bike) +
  geom_bar(mapping = aes(x = start_hour)) +
  facet_wrap(member_casual ~ start_day)
```

## Demonstrating map data

This is merely an example to demonstrate that while we might expect 
greater activity at the edges of the city in the morning, as commuters go into
work in the city the activity at the edges (suburbs) is rather weak. 
The real demand is clustered in the heart of the city.
We can speculate that perhaps customers are taking the "L" into the city and
using the bike for the final leg of their journey.

```{r plot_map, echo=FALSE}

ggplot(data = clean_bike) +
  geom_point(data = subset(clean_bike, start_hour = 8),
             mapping = aes(x = start_lat, y = start_lng, alpha = .05)) +
  facet_wrap(~member_casual)
```

## Ride Duration
Casual riders average a 71 minute ride while members average a 21 minute ride.
It is unlikely that casual riders are spending a hour on the bike commuting
everyday, therefore we can tentatively conclude that they are using the bike
primarily for recreation.
Members on the other hand average just 20 minutes, which is exactly right if
you were commuting by bike. 
Here I am drawing from personal experience as I commuted to work by bike for 
10 years while living in Baltimore. My commute covered 4 miles in about 
20 minutes depending on traffic lights. Which aligns perfectly with the data.

```{r avg_ride_duration}

ride_avg <- clean_bike %>% 
  group_by(member_casual, start_day, start_hour) %>% 
  summarize(ride_avg = mean(ride_duration) / 60)
```
## Ride Duration by Day
Here we graph the average ride time by day per member/casual.
Among casual riders We can see a fairly wide distribution of average 
ride time cresting somewhere in the middle of the day, while members stay
consistently in the 20 minute range day in and day out.
This would contradict our earlier theory that members are using the bike
as a form of recreation.

```{r plot_duration_by_day, echo=FALSE}

ride_time <- ggplot(data = ride_avg) +
  geom_point(mapping = aes(x=start_hour, y=ride_avg)) +
  geom_smooth(mapping = aes(x=start_hour, y=ride_avg),
              method = 'loess', formula = 'y ~ x') +
  scale_y_continuous(limits = c(0, 100)) +
  facet_grid(member_casual ~ start_day)
ride_time
```

# Conclusion
In our analysis we were tasked with answering three questions:

1. How do member and casual riders differ?

2. Why would casual riders buy into the membership program?

3. How should digital media be used to influence casual riders to become
members?


# How do member and casual riders differ.
Member riders are far more active across all hours of the day compared to 
casual riders.
Casual riders take longer rides, around an hour, most likely for recreation, 
which is supported by their increase usage on the weekend. 
Member riders take shorter trips, around 20 minutes, most likely as commuters.
However that fails to explain members continued strong usage on the weekend.
We can again theorize this is members running errands on the bike.
This observation is critical because it indicates members see the bike as
a method of transportation, a tool, **not a recreation**.

# Why would casual riders buy into the membership program?
To get casual members to buy into the membership program they must be convinced
that the bike is not (or is not *just*) a recreational activity but a means of
transportation; something they can use to get to work and run errands.

# How should digital media be used to influence casual riders to become members?
Digital media should focus on portraying the membership program as an effective
mode of transportation.
This campaign should target people who live or work in the downtown area as that
is where the overwhelming majority of use occurs.


# Follow up

*set timeline for follow up.

*set KPIs


# Further questions for analysis:
* It might be interesting to study the conversion rate of casual riders to
members as it would reveal if it is worthwhile pursuing marketing campaigns to 
draw in more casual riders who can later be converted to members.
* This data doesn't contain rider ids so we cannot say if there is a core of
riders who make up the majority of the activity or if it is evenly spread
amongst all users.
* demographic information about users would also allow more effective media
campaigns. Age, sex, education, etc.












